---
title: "Phenology Modelling"
author: "Nils Tinner"
date: "`r Sys.Date()`"
output: html_document
---

This excercise builds upon the model presented in "https://geco-bern.github.io/handfull_of_pixels/phenology_modelling.html". The first task is to think of ways to improve the model. The following are ideas that I came up with.

1) Add more training data
2) Add information about plant species
3) Use more parameters for model

--> 1+3 one is chosen to implement

```{r}
packages <- c("phenocamr","dplyr","ggplot2","lubridate","daymetr","leaflet","terra","MODISTools") 

source("../R/load_packages.R")
load_packages(packages)#loading packages

phenocamr::download_phenocam(
  site = "harvard$",
  veg_type = "DB",
  roi_id = "1000",
  daymet = TRUE,
  phenophase = TRUE,
  trim = 2022,
  out_dir = tempdir()
) #Downloading phenocam data

harvard_phenocam_data <- readr::read_csv(
  file.path(tempdir(), "harvard_DB_1000_3day.csv"),
  comment = "#"
) #phenocam data read in

# reading in harvard phenology only retaining
# spring (rising) phenology for the GCC 90th
# percentile time series (the default)
harvard_phenology <- readr::read_csv(
  file.path(
    tempdir(),
    "harvard_DB_1000_3day_transition_dates.csv"
  ),
  comment = "#"
) |>
  dplyr::filter(
    direction == "rising",
    gcc_value == "gcc_90"
  )

#Now we also load data from the bartlettir site. same as for harvard
phenocamr::download_phenocam(
  site = "bartlettir$",
  veg_type = "DB",
  roi_id = "1000",
  daymet = TRUE,
  phenophase = TRUE,
  trim = 2022,
  out_dir = tempdir()
)

bartel_phenocam_data <- readr::read_csv(
  file.path(tempdir(), "bartlettir_DB_1000_3day.csv"),
  comment = "#"
)

#we apply the same processing
bartel_phenology <- readr::read_csv(
  file.path(
    tempdir(),
    "bartlettir_DB_1000_3day_transition_dates.csv"
  ),
  comment = "#"
) |>
  dplyr::filter(
    direction == "rising",
    gcc_value == "gcc_90"
  )

```


```{r}
#harvard temp and also radiation since used to improve model
harvard_temp <- harvard_phenocam_data |>
  group_by(year) |>
  dplyr::mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.)/2
  ) |>
  dplyr::mutate(
    date = as.Date(date),
    gdd = cumsum(ifelse(tmean >= 5, tmean - 5, 0))
  ) |>
  dplyr::select(
    date,
    year,
    tmean,
    gdd,
    srad..W.m.2. #Now we added also solar radiation
  ) |>
  dplyr::ungroup()|>
  mutate(site = "Harvard") #we give a name to be able to identify the sites later on

#we do the same for bartel-site
bartel_temp <- bartel_phenocam_data |>
  group_by(year) |>
  dplyr::mutate(
    tmean = (tmax..deg.c. + tmin..deg.c.)/2
  ) |>
  dplyr::mutate(
    date = as.Date(date),
    gdd = cumsum(ifelse(tmean >= 5, tmean - 5, 0))
  ) |>
  dplyr::select(
    date,
    year,
    tmean,
    gdd,
    srad..W.m.2.
  ) |>
  dplyr::ungroup()|>
    mutate(site = "Bartel")
all_temp <- rbind(bartel_temp,harvard_temp) #then we combine the dataset

```
Now we process the phenology data
```{r}
#for Harvard
harvard_phenology <- harvard_phenology |>
  mutate(
    doy = as.numeric(format(as.Date(transition_25),"%j")),
    year = as.numeric(format(as.Date(transition_25),"%Y"))
  ) |>
  select(
    year,
    doy,
    transition_25,
    threshold_25
  )|>
    mutate(site = "Harvard")
#same for bartel
bartel_phenology <- bartel_phenology |>
  mutate(
    doy = as.numeric(format(as.Date(transition_25),"%j")),
    year = as.numeric(format(as.Date(transition_25),"%Y"))
  ) |>
  select(
    year,
    doy,
    transition_25,
    threshold_25
  )|>
    mutate(site = "Bartel")


all_pheno <- rbind(bartel_phenology,harvard_phenology) #again we combine

```


```{r}
gdd_model <- function(data_met, par) {
  # split out parameters from a simple
  # vector of parameter values
  temp_threshold <- par[1]
  gdd_crit <- par[2]
  solar_impact <- par[3]

  #taking apart the meteodata fed to the model
  temp <- data_met[seq(from = 1, to = length(data_met)/2,by = 1)]
  sol <- data_met[seq(from = length(data_met)/2+1, to = length(data_met),by = 1)]


  #using temp for growing degree days
  gdd <- cumsum(ifelse(temp > temp_threshold, temp - temp_threshold, 0))
  gdd <- gdd + (sol-mean(sol))/sd(sol)*solar_impact #A z standardized impact of the precipitation on gdd...



  doy <- unlist(which(gdd >= gdd_crit)[1]) #doy

  return(doy)
}

```

```{r}
rmse_gdd <- function(par, data) {

  # split out data
  drivers <- data$drivers
  validation <- data$validation



  # calculate phenology predictions
  # and put in a data frame
  predictions <- drivers |>
    group_by(year,site) |> #now also sites
    summarise(
      predictions = gdd_model(
        data_met = c(tmean,srad..W.m.2.),#temp and solar rad
        par = par
      ),
      .groups = 'drop'
    )|>
    ungroup()

  predictions <- left_join(predictions, validation, by = c("year","site"))

  rmse <- predictions |>
    summarise(
      rmse = sqrt(mean((predictions - doy)^2, na.rm = TRUE))
    ) |>
    pull(rmse)

  # return rmse value
  return(rmse)
}
```


```{r}
# starting model parameters
par = c(0, 130,1)

# limits to the parameter space
lower <- c(-10,0,0.000001)
upper <- c(45,500,10000)

# data needs to be provided in a consistent
# single data file, a nested data structure
# will therefore accept non standard data formats
data <- list(
  drivers = all_temp,
  validation = all_pheno
)

# optimize the model parameters
optim_par = GenSA::GenSA(
  par = par,
  fn = rmse_gdd,
  lower = lower,
  upper = upper,
  control = list(
    max.call = 4000
  ),
  data = data
)$par



```



```{r}
#make predictions
predictions <- all_temp |>
  group_by(year,site) |>
  summarize(
    prediction = gdd_model(
      data_met = c(tmean,srad..W.m.2.),
      par = optim_par
    )
  )


# join predicted with observed data
validation <- left_join(predictions, all_pheno)
#plot with respect to sites
ggplot2::ggplot(validation) +
  geom_smooth(
    aes(
      doy,
      prediction,
    ),
    colour = "grey25",
    method = "lm"
  ) +
  geom_point(
    aes(
      doy,
      prediction,
      col = site
    )
  ) +
  geom_abline(
    intercept=0,
    slope=1,
    linetype="dotted"
  ) +
  labs(
    x = "Observed leaf-out date (DOY)",
    y = "Predicted leaf-out date (DOY)"
  ) +
  theme_bw()


```



```{r}


# Download daily data, this may take a wile, other data is used later
# daymetr::download_daymet_tiles(
#   tiles = 11935,
#   start = 2008,
#   end = 2012,
#   param = c("tmin","tmax","srad"),
#   path = paste0(here::here(), "/data-raw/"),
#   silent = TRUE
#   )

# calculate the daily mean values
r <- daymetr::daymet_grid_tmean(
  path = paste0(here::here(), "/data-raw/"),
  product = 11935,
  year = 2012,
  internal = TRUE
)#temperature

r_2 <- terra::rast(paste0(here::here(), "/data-raw/","srad_2012_11935.nc")) #radiation



# reproject to lat lon
r <- terra::project(
  r,
  "+init=epsg:4326"
)


# subset to first 180 days
ma_nh_temp <- terra::subset(
  r,
  1:180
)

# reproject to lat lon also for rad
r_2 <- terra::project(
  r_2,
  "+init=epsg:4326"
)


# subset to first 180 days
ma_nh_sol <- terra::subset(
  r_2,
  1:180
)
```

```{r}


predicted_phenology <- terra::app(
  c(ma_nh_temp,ma_nh_sol),
  fun = gdd_model,
  par = optim_par
)


```



```{r}



# set te colour scale manually
pal <- colorNumeric(
  "magma",
  values(predicted_phenology),
  na.color = "transparent"
)

# build the leaflet map
# using ESRI tile servers
# and the loaded demo raster
leaflet() |>
  addProviderTiles(providers$Esri.WorldImagery, group = "World Imagery") |>
  addProviderTiles(providers$Esri.WorldTopoMap, group = "World Topo") |>
  addRasterImage(
    predicted_phenology,
    colors = pal,
    opacity = 0.8,
    group = "Phenology model results"
  ) |>
  addLayersControl(
    baseGroups = c("World Imagery","World Topo"),
    position = "topleft",
    options = layersControlOptions(collapsed = FALSE),
    overlayGroups = c("Phenology model results")
  ) |>
  addLegend(
    pal = pal,
    values = terra::values(predicted_phenology),
    title = "DOY")

```

```{r}



# download and save phenology data
phenology <- MODISTools::mt_subset(
  product = "MCD12Q2",
  lat = 43.152662,
  lon =  -71.382337,
  band = "Greenup.Num_Modes_01",
  start = "2012-01-01",
  end = "2012-12-31",
  km_lr = 50,
  km_ab = 50,
  internal = TRUE,
  progress = TRUE
)

phenology <- phenology |>
  mutate(
    value = as.numeric(format(as.Date("1970-01-01") + value, "%j")),#mutate dates...
    value = ifelse (value < 200, value, NA)) #data quality control
#now raster
phenology_raster <- MODISTools::mt_to_terra(
  phenology,
  reproject = TRUE
)

#now processing: cropping to each others extend...
predicted_phenology <- terra::crop(
  x = predicted_phenology,
  y = phenology_raster
)

phenology_raster <- terra::crop(
  x = phenology_raster,
  y = predicted_phenology
)

#and adjusting resolution
phenology_raster <- terra::resample(
  x = phenology_raster,
  y = predicted_phenology,
  method = "average"
)

predicted_phenology <- terra::mask(
  predicted_phenology,
  is.na(phenology_raster),
  maskvalues = TRUE
)

 ggplot() +
  tidyterra::geom_spatraster(data = phenology_raster) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "DOY"
    ) +
  theme_bw()
 
  ggplot() +
  tidyterra::geom_spatraster(data = predicted_phenology) +
  scale_fill_viridis_c(
    na.value = NA,
    name = "DOY"
    ) +
  theme_bw()
 
#resulting in these two maps:
```
  
  
```{r}
  
  #Now we compare for the year 2012
  predicted_df <- as.vector(predicted_phenology)
phenology_df <- as.vector(phenology_raster)
sct_df <- data.frame(
  doy_p = predicted_df,
  doy_m = phenology_df
  )

ggplot(
  data = sct_df,
      aes(
      doy_p,
      doy_m
    )
  ) +
  geom_point() +
  scale_fill_viridis_c(trans="log10") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = "white",
    lty = 2
  ) +
  labs(
    x = "predicted maturity (DOY)",
    y = "MODIS vegetation maturity (DOY)"
  ) +
  theme_bw()
```
Not looking very promising, lets generalize this to the years 2008-2012
```{r}
sct_df <- data.frame(
  doy_p = NULL,
  doy_m = NULL,
  year = NULL
  )
for (temp_year in c(2008:2012)) {
  print(temp_year)
  print(str(sct_df))

# calculate the daily mean values
r <- daymetr::daymet_grid_tmean(
  path = paste0(here::here(), "/data-raw/"),
  product = 11935,
  year = temp_year,
  internal = TRUE
)
r_2 <- terra::rast(paste0(here::here(), "/data-raw/","srad_",temp_year,"_11935.nc"))



# reproject to lat lon
r <- terra::project(
  r,
  "+init=epsg:4326"
)


# subset to first 180 days
ma_nh_temp <- terra::subset(
  r,
  1:180
)

# reproject to lat lon
r_2 <- terra::project(
  r_2,
  "+init=epsg:4326"
)


# subset to first 180 days
ma_nh_sol <- terra::subset(
  r_2,
  1:180
)



predicted_phenology <- terra::app(
  c(ma_nh_temp,ma_nh_sol),
  fun = gdd_model,
  par = optim_par
)
# download and save phenology data
phenology <- MODISTools::mt_subset(
  product = "MCD12Q2",
  lat = 43.152662,
  lon =  -71.382337,
  band = "Greenup.Num_Modes_01",
  start = paste0(temp_year,"-01-01"),
  end = paste0(temp_year,"-12-31"),
  km_lr = 50,
  km_ab = 50,
  internal = TRUE,
  progress = TRUE
)

phenology <- phenology |>
  mutate(
    value = as.numeric(format(as.Date("1970-01-01") + value, "%j")),
    value = ifelse (value < 200, value, NA))

phenology_raster <- MODISTools::mt_to_terra(
  phenology,
  reproject = TRUE
)


predicted_phenology <- terra::crop(
  x = predicted_phenology,
  y = phenology_raster
)

phenology_raster <- terra::crop(
  x = phenology_raster,
  y = predicted_phenology
)


phenology_raster <- terra::resample(
  x = phenology_raster,
  y = predicted_phenology,
  method = "average"
)

predicted_phenology <- terra::mask(
  predicted_phenology,
  is.na(phenology_raster),
  maskvalues = TRUE
)
print(str(data.frame(
  doy_p = predicted_df,
  doy_m = phenology_df,
  year = temp_year
  )))

 predicted_df <- as.vector(predicted_phenology)
phenology_df <- as.vector(phenology_raster)
sct_df <- rbind(sct_df,data.frame(
  doy_p = predicted_df,
  doy_m = phenology_df,
  year = temp_year
  ))

}









  predicted_df <- as.vector(predicted_phenology)
phenology_df <- as.vector(phenology_raster)
sct_df <- data.frame(
  doy_p = predicted_df,
  doy_m = phenology_df
  )

ggplot(
  data = sct_df,
      aes(
      doy_p,
      doy_m,
      col = year
    )
  ) +
  geom_point() +
  scale_fill_viridis_c(trans="log10") +
  geom_smooth(
    method = "lm",
    se = FALSE,
    colour = "white",
    lty = 2
  ) +
  labs(
    x = "predicted maturity (DOY)",
    y = "MODIS vegetation maturity (DOY)"
  ) +
  theme_bw()
```
Alright, looks much better...
